---
title: "WORLD"
author:
  name: "Amir Ahmed Seid"
  email: "amirseid1997@gmail.com"
format:
  html:
    theme: flatly
    toc: true
    toc-floating: true
execute: 
  warning: false
  message: false
editor-options:
  chunk-output-type: console
---

# Visualizing the COVID-19 Data Set

In this session, we will perform data wrangling, exploratory data analysis, and visualization. We got this dataset from Kaggle. You can access it through this link.

This session is not a report but rather a go through over what my thought process looked like when navigating the data.

## Prerequistes

We need to download certain libraries for our analysis.

```{r}
library(tidyverse) # data cleaning
library(scales) # editting labels in our plot
library(patchwork) # plot many graphs in a single canva
library(gganimate) # to create animations
library(gifski) # to convert animations to gif
library(ggthemes) # to get a map theme
library(rnaturalearth) #We can use maps, but the best package for geospatial analysis
library(sf) # work with sf geometry to create maps
library(maps) # us map data, in case.
library(magick) # to read saved gif files.

```

### Load the CSV file

The first step when working with a data set is to load it into our R session and get an overview of it.

```{r}
path <- "tested_worldwide.csv"
COVID_raw <- read.csv(path)
```

Let's gather some general information about the type of data set we are dealing with.

```{r}
str(COVID_raw)
```

We have `27641` observations and `12` variables. Many of our variables are epidemiological outcomes of COVID-19. Since our data values are unlikely to have causality with one another, this session is more focused on reporting and summarizing.

We can answer various questions using the dataset. For instance,

-   We can plot visualizations to show changes in recovery, mortality, incidence, etc.
-   We can create an incidence chart across a specific country within the period.
-   We can develop a time series analysis to forecast the future.

### Exploring the data

We can't know what we are able to do with our data before exploring it. So, let's look what our data looks like.

```{r}
df <- COVID_raw
```

The data type for `Date` should be a datetime type of data instead of a string.

```{r}
df <- df %>% 
    mutate(Date = as.Date(Date, format = "%Y-%m-%d"))
str(df)
```

Let's see how many NA values we have in each column.

```{r}
na_count <- function(dataframe){
  Names_cols <- colnames(dataframe)
  a <- sapply(Names_cols,
          function(c){
           round(mean(is.na(dataframe[, grepl(c, names(df))])), 2)
         }
         )
  my_df <- data.frame(
    Columns = attr(a, 'names'),
    na_prcnt = a,
    row.names = NULL
    )
  return(my_df)
}

na_count(df)
```

We can also use the `colSums()` function to see na_vals in our data frame.

Some variables have too many missing values that we cannot perform any task as a data analyst, except to present them in visuals. Even then, they don't give viable information about the phenomenon. `hospitalized`, `active`, and `hospitalizedCurr` have too many missing values. Therefore, it is safer to remove them.

This cleaning is done after exploring our data. This is done to write clean data cleaning code with clear steps.

Let us see the distribution of our observations.

```{r}
fig1 <- df %>% 
  mutate(Country_Region = fct_lump(Country_Region, n = 15)) %>% 
  count(Country_Region, sort = T) %>% 
  mutate(Country_Region = fct_reorder(Country_Region, n)) %>% 
  ggplot(aes(Country_Region, n, fill = Country_Region)) + 
  geom_col(show.legend = FALSE)+
  coord_flip()+
  theme_test()+
  labs(
    title = "Data registration count distribution across Countries",
    x = "Countries",
    y = "Counts"
  )+
  theme(plot.title = element_text(
    size = 8,
    hjust = 0.5,
    vjust = 1,
    margin = margin(b = 10) 
  ))

fig2 <- df %>% 
  mutate(Country_Region = fct_lump(Country_Region, n = 15)) %>% 
  count(Country_Region, sort = T) %>% 
  mutate(Country_Region = fct_reorder(Country_Region, n)) %>% 
  ggplot(aes(n)) +
  geom_boxplot()+
  labs(
    title = "Distribution of Registry Counts Across Countries",
    x = "Count"
  )+
  theme(plot.title = element_text(
    size = 8,
    hjust = 0.5,
    vjust = 1,
    margin = margin(b = 10)
  ))+
  theme_test()
```

```{r, echo=FALSE}
fig1 + fig2
```

Our visualization tells us that our data has fewer data points for countries outside the USA, Australia, and Canada. Hence, it is wise to make our time series model, for instance, about the incidence of COVID-19 in these countries.

```{r}
df <- df %>% 
  filter((!(is.na(total_tested) | is.na(positive))) & total_tested >= 25) %>%
  mutate(Country_Region = 
           if_else(Country_Region %in% c("Turkey", "United States", "Australia", "Canada"),
                   Country_Region, "Others"),
         incidence_rate = round((positive / total_tested), 2))
```

```{r}
df %>% 
  mutate(
    Date = floor_date(Date, "week")
  ) %>% 
  group_by(Date, Country_Region) %>%
  summarise(
    incidence_rate = mean(incidence_rate, na.rm = TRUE),
    Count = n()
  ) %>%
  filter(!is.na(incidence_rate)) %>% 
  ggplot(aes(Date, incidence_rate, colour = Country_Region))+
  geom_line(linewidth = 1)+
  scale_y_continuous(labels = scales::percent)+
  theme_test()+
  labs(
    y = "Incidence rate",
    title = "COVID-19 Positivity Rate"
  )+
  theme(plot.title = element_text(
    size = 12,
    vjust = 1,
    margin = margin(b = 10)
  ))
```

Let us plot the **death/ positive** people ratio for our countries.

```{r}
df %>% 
  mutate(
    Date = floor_date(Date, "week")
  ) %>% 
  group_by(Date, Country_Region) %>%
  summarise(
    CFR = round(mean((death/ positive), na.rm = T), 2),
    Count = n(),
    .groups = "drop"
  ) %>% 
  filter(!is.na(CFR)) %>% 
  ggplot(aes(Date, CFR, colour = Country_Region))+
  geom_line(linewidth = 1)+
  scale_y_continuous(labels = scales::percent)+
  theme_test()+
  labs(
    y = "Incidence rate",
    title = "COVID-19 Positivity Rate"
  )+
  theme(plot.title = element_text(
    size = 12,
    vjust = 1,
    margin = margin(b = 10)
  ))
```

Both the PR and CFR plot graphs show us that Australia's dataset has a data entry error. The CFR value is 1/∞ while the positivity value is ∞. Australia's data has noise. Therefore, we need to remove Australia from our analysis.

At this point, we have understood our data. But, we need to have a clean script so people can understand our analysis when we share it, and call our clean dataframe `COVID`

```{r}
COVID <- COVID_raw %>% 
  filter(positive != 0, death != 0,
         !is.na(total_tested), !is.na(positive), total_tested >= 25, Country_Region != "Australia") %>% 
        
        #edit existing columns
        # Time date format
        mutate(Date = as.Date(Date, format = "%Y-%m-%d"),

              #Remove many categorical variables
              Country_Region = 
                 if_else(Country_Region %in% c("Turkey", "United States", "Canada"),
                         Country_Region, "Others"),
              
              #Add new parameters
              # filter before these: positive and death can't be 0.
              positivity_rate = round((positive / total_tested), 2),
              CFR = round((death/ positive), 2)
               ) %>% 
          # Wrong values
          filter(CFR < 1 & CFR >= 0) %>% 
          #remove columns with significant NA values
          select(-c(hospitalized, hospitalizedCurr)) %>% 
          rename(Country = Country_Region)
```

## Visualization

-   Smoothed faceted time graph for incidence rate.
-   Smoothed faceted time graph for CFR(Case Fatality Rate)
-   Map animation for both incidence rate and CFR for the USA

Let us first create a clean environment ready to perform our upcoming graphs, maps, and animations.

```{r}
COVID_4graphs <- COVID %>% 
  mutate(Date = floor_date(Date, "day")) %>% 
  group_by(Date, Country) %>% 
  summarise(
    positivity_rate = mean(positivity_rate, na.rm = T),
    CFR = mean(CFR, na.rm = T),
    .groups = "drop"
  ) 

COVID_4graphs %>% 
ggplot(aes(Date, positivity_rate, color = Country)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent)+
  facet_wrap(~Country, scales = "free_y")+
  theme_test()+
  labs(
    title = "Positivity Rate of COVID-19 tests in the World",
    x = "",
    y = ""
  )+
  theme(legend.position = 'none')
```

If you have noticed in the code, we have included this line `mutate(Date = floor_date(Date, "day"))`, which is from the lubridate package. Although it looks unnecessary, we added it so that we can adjust and observe the trend over weeks or months when needed.

We can reuse the same formula, with minimal adjustments, to plot CFR.

```{r}
COVID_4graphs %>%  
  ggplot(aes(Date, CFR, color = Country)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent)+
  facet_wrap(~Country, scales = "free_y")+
  theme_test()+
  labs(
    title = "Case Fatality Rate(CFR) of COVID-19 tests in the World",
    x = "",
    y = ""
  )+
  theme(legend.position = 'none')
```

We can observe that generally, there is a decrease in CFR across our timeline. On the other hand, we have noticed that there is a noise in Turkiye's observations. Usually, a straight line in a graph shows that there are missing values, and ggplot2 is trying to forward fill the missing points.

### Animation

-   We will create an animated plot of the USA.

We will create our animations using the gganimate library.

```{r}
COVID_4graphs %>% 
ggplot(aes(Date, positivity_rate, color = Country)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent)+
  theme_test()+
  labs(
    title = "Positivity Rate of COVID-19 tests in the World",
    caption =  "Date:{frame_along}",
    x = "",
    y = ""
  )+
  transition_reveal(Date)
```

```{r}
COVID_4graphs %>%  
  ggplot(aes(Date, CFR, color = Country)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent)+
  theme_test()+
  labs(
    title = "Case Fatality Rate(CFR) of COVID-19 tests in the World",
    caption =  "Date:{frame_along}",
    x = "",
    y = ""
  )+
  theme(axis.title = element_text(
      hjust = 1,
      vjust = 2.5
  ))+
  transition_reveal(Date)
```

### Map

Let us find the geospatial data we need to create the map of the USA. Once we get it, we will combine it with our `COVID_USA` dataframe.

```{r}
COVID_USA <- COVID %>%
  filter(Country == "United States") %>% 
  rename(State = Province_State)

COVID_USA %>% 
  count(State, wt = daily_tested) %>% 
  arrange(desc(n)) %>% 
  head()
```

```{r, warning=FALSE}
USA <- ne_states(country = "United States of America",
                 returnclass = "sf")

USA <- USA %>% 
  select(name_en, name_tr, geometry) %>% 
  rename(State = name_en)

COVID_USA <- COVID_USA %>% left_join(USA, by = "State") %>%
                  st_as_sf()
```

Since we are creating an animation, we need to reduce our observations to save computing power. We will group the data per month.

```{r}
COVID_USA <- COVID_USA %>% 
  mutate(
    Date = floor_date(Date, "months"))
```

Let us prepare a clean dataframe, for the last time here, ready for our animation.

We want to create an animation of the Case Fatality Rate. So, we need the mean value of each state's monthly CFR and the sf_geometry of each state in our dataframe

```{r}
COVID_USA_4map <- COVID_USA %>% 
  filter(State != "All States") %>% 
  group_by(Date, State, geometry) %>% 
  summarise(
    CFR = round(mean(CFR, na.rm = T), 2),
    .groups = "drop"
  ) 
  
COVID_USA_4map <- COVID_USA_4map %>% 
  distinct(Date, State, .keep_all = TRUE) 
```

If we try to plot an animated plot, it will crash for a few reasons. One reason is that we don't have enough data for the first two months to plot. If you need a detailed explanation about how **gganimate** works, please pay a visit to your nearest data analyst. In my defense, this is my table:

```{r}
COVID_USA_4map %>% 
  count(Date)
```

```{r eval=FALSE, message=FALSE, warning=FALSE}

COVID_USA_4map <- COVID_USA_4map %>% 
  filter(Date >= "2020-04-01")

p <- COVID_USA_4map %>%
  ggplot() +
  geom_sf(aes(fill = CFR)) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = median(COVID_USA$CFR, na.rm = TRUE),
    na.value = "grey80"
  ) +
  coord_sf(xlim = c(-170, -65), ylim = c(25, 70)) +
  theme_map() +
  labs(
    title = "COVID-19 Case Fatality Rate by US State\nMonth: {format(frame_time, '%B %Y')}",
    fill = "CFR"
  ) +
  transition_time(Date) +
  ease_aes('linear')

animated_gif <- animate(p, renderer = gifski_renderer(), width = 800, height = 400, fps = 1, duration = 8)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
gif <- image_read("animated.gif")

image_animate(gif, fps = 2, delay = 40)
```


In today's analysis, we have seen:

-   **Data Wrangling**: we cleaned and prepared the raw data, handled the anomailities in our data using tidyverse
-   **Exploratory Data Analysis**: We identified patterns in CFR and Positivity Rate using tidyverse
-   **Visualization**: we plotted ggplots, maps, and animated graphs using gganimate.
-   **Spatial Analysis** using the packages maps, sf, and ggthemes.

We will see how to create a time series analysis using ARIMA.

NB: This webpage is a work in progress and is not the final version. The analysis here is for educational and exploratory purposes. So, it should be treated as such and shouldn't be used as a qualified medical, public health, or epidemiological finding. The conclusions reached here are limited by the data available and the statistical methods opted to be followed.

Created by - Amir Seid

Medical Student

Quantitative Research

Data Science

Application of ML in Healthcare.
